On particular diameter bounds for integral point sets in higher dimensions \footnote{ This work was carried out at Voronezh State University and supported by Russian Science Foundation grant 19-11-00197. }

### 2021-11-08 AvdMomZvol_rails/Article.tex
[AMZr] Probably the last example of 4-4-rails
[AMZr] The bunch of final edits

\paragraph{Absctract.}
	An integral point set $\mathcal{P}$ is a set of $n$ points in
	the $m$-dimensional Euclidean space $\mathbb{R}^{m}$ with pairwise
	integral distances,
	such that $\mathcal{P}$ is not contained in a $(m-1)$-dimensional hyperplane.
	In the present paper we present some classes of planar integral point sets ($m=2$);
	we mostly focus on the subsets of a union of two straight lines:
	facher, rails, scissors and pyramid sets.
	We use rails and pyramid sets to construct lower bounds for a diameter
	of integral point sets in higher dimensions.

@@

@@

\begin{dfn}
	The function $d(m, n)$ is the minimal possible diameter of
	the integral point set $\mathcal{P}$ of $n$ points in
	$m$-dimensional Euclidean space $\mathbb{R}^{m}$.
\end{dfn}

@@

%For a bit more sophisticated configurations, another functions are presented~\cite{kurz2008minimum,kreisel2008heptagon}.
%
%\begin{dfn}
%	The function $\overline{d}(m, n)$ is the minimum possible diameter of
%	the integral point set $\mathcal{P}$ of $n$ points in
%	$m$-dimensional Euclidean space $\mathbb{R}^{m}$
%	such that no $m+1$ points are located on an $(m-1)$-dimensional hyperplane.
%\end{dfn}

%\begin{dfn}
%	The function $\dot{d}(m, n)$ is the minimum possible diameter of
%	the integral point set $\mathcal{P}$ of $n$ points in
%	$m$-dimensional Euclidean space $\mathbb{R}^{m}$
%	such that no three points are located on a straight line
%	and no four points are located on a circle.
%\end{dfn}

A long list of known bounds for $d(m, n)$ and some exact values
can be found in~\cite[Theorem 1]{kurz2008bounds} or in~\cite{our-vmmsh-2018}.
Below we will discuss the following ones presented at~\cite{kurz2008bounds}:
\begin{align}
	d(m, 2m + 1) \leq 8\\
	d(m, 2m + 2) \leq 13 \label{eq:d_m_2m+2}\\
	d(m, 3m) \leq 109
\end{align}
and the following theorem \cite[Theorem 2.1]{kurz2008bounds}.

@@

In this paper we further study the various features of IPS.
In Section~\ref{sec:classif}, we discuss the classification of planar integral points sets;
in Sections~\ref{sec:bounds_rails}~and~\ref{sec:bounds_pyramid}, we present some bounds for $d(m,n)$ based on planar integral point sets of special types
and provide some general constructions of such bounds.

@@

\subsection{Integral point sets situated in two straight lines}

@@

Non-facher integral point sets situated in two straight lines
can be further classificated in the following three cases:

\begin{dfn}
	A planar integral point set situated in two parallel straight lines
	is called a \textit{rails} set.
\end{dfn}

Among the rails sets, sets with 2 points on one line and all the other on another line prevail.
In Section~\ref{sec:bounds_rails} below, some examples of such sets are discussed;
they include an example of rails sets with 2 points on one line and 38 point on the other line
(Fig.~\ref{picture_40.png})
and
an example of rails set with 4 points on one line and 4 points on the other
(Fig.~\ref{picture.png}).
We have no examples of rails set with 4 points on one line and 5 points on the other;
and we have no examples of rails sets with 3 points on one line and 9 points on the other.

@@

On Fig.~\ref{11_2432_255255_a0a1b7f805cfc93080b4608505a0f45d_rails_3.png} we show the following rails set with 3 points on one line and 8 points on the other:
\begin{multline*}
\mathcal{P}_{3,8}=\sqrt{255255}/2*\{
	( 1767 ; -3);
	( 2791 ; -3);
	( 4071 ; -3);
\\
	( -306 ; 0);
	( 0 ; 0);
	( 1798 ; 0);
	( 2304 ; 0);
	( 2760 ; 0);
	( 3534 ; 0);
	( 4040 ; 0);
	( 4558 ; 0)
\}
\end{multline*}
It's notable that $\operatorname{char} \mathcal{P}_{3,8}=255255 = 3 \cdot 5 \cdot 7 \cdot 11 \cdot 13 \cdot 17$
is a production of the first 6 odd primes,
and the distances between the three points on the bottom line are $512=2^{9}$, $640=2^7 \cdot 5$ and
$512+640 = 1152 = 2^7 \cdot 3^2$.

@@

\begin{dfn}
	A planar integral point set situated in two perpendicular straight lines
	is called a \textit{cross} set.
\end{dfn}
Every cross set has characteristic 1;
in~\cite{antonov2008maximal}, cross sets with only 2 points out of one of the lines are called \textit{crabs}.

\begin{dfn}
	A planar integral point set situated in two straight lines
	that are not parallel nor perpendicular,
	is called a \textit{scissors} set.
\end{dfn}

@@

It still remain unknown whether a scissors set can have another axis of symmetry;
on Fig.~\ref{15_3364_91_30e991f82a54fb9caeca5b6a6776b981_X.png} the following scissors set with central symmetry is shown:
\begin{multline*}
	\sqrt{91}/1*\{
	( 0 ; 0);
	( 792 ; 120);
	( -792 ; -120);
\\
	( \pm256 ; 0);
	( \pm646 ; 0);
	( \pm693 ; 0);
	( \pm1152 ; 0);
	( \pm1242 ; 0);
	( \pm1682 ; 0)
\}
\end{multline*}

@@

There is still no general construction for a rails or scissors PIPS of arbitrary cardinality.
For rails PIPSs, we can conjecture that there exists a set of arbitrary cardinality, with 2 points on one line
and all the rest on the other;
on the other hand, we still have not found any rails PIPSs with 4 and 5 points on the lines.

@@

\section{Acknowledgements}
The author thanks Dr. A.S. Usachev for his help in polishing the text of the paper.

>>>>

\paragraph{Absctract.}
	An integral point set $\mathcal{P}$ is a set of $n$ points in
	the $m$-dimensional Euclidean space $\mathbb{R}^{m}$ with pairwise
	integral distances,
	such that $\mathcal{P}$ is not contained in a $(m-1)$-dimensional hyperplane.
	In the present paper we discuss some classes of planar integral point sets ($m=2$);
	we mostly focus on the subsets of a union of two straight lines:
	facher, rails, scissors and pyramid sets.
	We use rails and pyramid sets to construct lower bounds for a diameter
	of integral point sets in higher dimensions.

@@

The study of integral point sets belongs to the classical analysis.
It is motivated by both pure theoretical aspects~\cite{ascher2020erdos,solymosi2010question}
and applications~\cite{kurz2007enumeration}.

@@

\begin{dfn}
	The function $d(m, n)$ is the minimal possible diameter of
	an integral point set $\mathcal{P}$ of $n$ points in
	$m$-dimensional Euclidean space $\mathbb{R}^{m}$.
\end{dfn}

@@

A long list of known bounds for $d(m, n)$ and some exact values
can be found in~\cite[Theorem 1]{kurz2008bounds} or in~\cite{our-vmmsh-2018}.
Below we will discuss the following ones presented at~\cite{kurz2008bounds}:
\begin{align}
	d(m, 2m + 1) \leq 8  \label{eq:d_m_2m+1}\\
	d(m, 2m + 2) \leq 13 \label{eq:d_m_2m+2}\\
	d(m, 3m) \leq 109    \label{eq:d_m_3m}
\end{align}
and the following theorem \cite[Theorem 2.1]{kurz2008bounds}.

@@

In this paper we further study the various features of IPS.
In Section~\ref{sec:classif}, we discuss the classification of planar integral points sets;
in Sections~\ref{sec:bounds_rails}~and~\ref{sec:bounds_pyramid}, we present some upper bounds for $d(m,n)$
based on planar integral point sets of special types
and provide some general constructions of such bounds.
All the bounds, like~\eqref{eq:d_m_2m+1}--\eqref{eq:d_m_3m}, are of the form $d(m, km+p)\leq c_{k,p}$.

@@

\subsection{Integral point sets situated on two straight lines}

@@

Non-facher integral point sets situated on two straight lines
can be further classificated in the following three cases:

\begin{dfn}
	A planar integral point set situated on two parallel straight lines
	is called a \textit{rails} set.
\end{dfn}

Among the rails sets, sets with 2 points on one line and all the other on another line prevail.
In Section~\ref{sec:bounds_rails} below, some examples of such sets are discussed;
they include an example of rails sets with 2 points on one line and 38 point on the other line
(Fig.~\ref{picture_40.png})
and
an example of rails set with 4 points on one line and 4 points on the other
(Fig.~\ref{picture.png}).
We still do not know whether there are rails set with 4 points on one line and 5 points on the other;
and the same for rails sets with 3 points on one line and 9 points on the other.

@@

On Fig.~\ref{11_2432_255255_a0a1b7f805cfc93080b4608505a0f45d_rails_3.png} we show the following rails set with 3 points on one line and 8 points on the other:
\begin{multline*}
\mathcal{P}_{3,8}=\sqrt{255255}/2*\{
	( 1767 ; -3);
	( 2791 ; -3);
	( 4071 ; -3);
\\
	( -306 ; 0);
	( 0 ; 0);
	( 1798 ; 0);
	( 2304 ; 0);
	( 2760 ; 0);
	( 3534 ; 0);
	( 4040 ; 0);
	( 4558 ; 0)
\}
\end{multline*}
It's notable that $\operatorname{char} \mathcal{P}_{3,8}=255255 = 3 \cdot 5 \cdot 7 \cdot 11 \cdot 13 \cdot 17$
is a product of the first 6 odd primes,
and the distances between the three points on the lower line are $512=2^{9}$, $640=2^7 \cdot 5$ and
$512+640 = 1152 = 2^7 \cdot 3^2$.

@@

Rails sets with 4 and 4 points on the lines are somewhat rather common;
an example of such a set is shown on Fig.~\ref{8_1111_770_5666c83a03a80e1d3f8f8e8186a145f6__rails_4.png}:
\begin{equation*}
	\sqrt{770}/1*\{
		( 0 ; 0);
		( 380 ; 0);
		( 731 ; 0);
		( 1111 ; 0);
		( 354 ; -12);
		( 377 ; -12);
		( 734 ; -12);
		( 757 ; -12)%;
	\}
\end{equation*}

\begin{figure}[h!]
	\center{\includegraphics[width=0.9\linewidth]{8_1111_770_5666c83a03a80e1d3f8f8e8186a145f6__rails_4.png}}
	\parbox{0.9\linewidth}{\caption{PIPS of cardinality 8 and diameter 1111}
	\label{8_1111_770_5666c83a03a80e1d3f8f8e8186a145f6__rails_4.png}}
\end{figure}

\begin{dfn}
	A planar integral point set situated on two perpendicular straight lines
	is called a \textit{cross} set.
\end{dfn}
Every cross set has characteristic 1;
in~\cite{antonov2008maximal}, cross sets with only 2 points out of one of the lines that have two axes of symmetry are called \textit{crabs}.

\begin{dfn}
	A planar integral point set situated on two straight lines
	that are not parallel nor perpendicular,
	is called a \textit{scissors} set.
\end{dfn}

@@

It still remains unknown whether a scissors set can have another axis of symmetry;
on Fig.~\ref{15_3364_91_30e991f82a54fb9caeca5b6a6776b981_X.png} the following scissors set with central symmetry is shown:
\begin{multline*}
	\sqrt{91}/1*\{
	( 0 ; 0);
	( 792 ; 120);
	( -792 ; -120);
\\
	( \pm256 ; 0);
	( \pm646 ; 0);
	( \pm693 ; 0);
	( \pm1152 ; 0);
	( \pm1242 ; 0);
	( \pm1682 ; 0)
\}
\end{multline*}

@@

There is still no general construction for a rails or scissors PIPS of arbitrary cardinality.
For rails PIPSs, we can conjecture that there exists a set of arbitrary cardinality, with 2 points on one line
and all the rest on the other;
on the other hand, it is still unknown whether there are any rails PIPSs with 4 and 5 points on the lines.

@@

\section{Acknowledgements}
The authors thank A.S. Usachev for his help in polishing the text of the paper.
